Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-12-24 paper/paper.tex
_Immunity_ formatting

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{nameref}

@@

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

@@

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\LARGE
\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity}}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Alistair B. Russell\textsuperscript{1},
Jacob R. Kowalsky\textsuperscript{2},
Jesse D. Bloom\textsuperscript{1,2,3*}
\\
\bigskip
{\small
$^1$Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center
\\
$^2$Department of Genome Sciences, University of Washington
\\
$^3$Howard Hughes Medical Institute
\\
Seattle, WA  98109, USA
\\
*Correspondence: jbloom@fredhutch.org
}
\bigskip

@@

\section*{Summary}
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.

@@

\section*{Results}

@@

\section*{Discussion}
We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\citep{saikia2018simultaneous, zanini2018virus}.
To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \citep{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.

@@

\small
\subsection*{Acknowledgments}
We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
JDB is an Investigator of the Howard Hughes Medical Institute.

\subsection*{Author Contributions}
ABR and JDB designed the study.
ABR performed the experiments with assistance from JRK.
ABR and JDB computationally analyzed the data and wrote the paper.

\subsection*{Declaration of Interests}
The authors declare no competing interests.

@@

\section*{Methods}

@@

\section*{Supporting information}

>>>>

@@

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\definecolor{linkblue}{RGB}{0,76,153}
\usepackage[colorlinks=true,linkcolor=linkblue,citecolor=linkblue,urlcolor=linkblue]{hyperref}
\usepackage{nameref}

@@

% Text layout
%\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

@@

% section header formats
\usepackage{sectsty}
\definecolor{darkred}{RGB}{140,0,0}
\sectionfont{\normalsize \color{darkred}}
\subsectionfont{\normalsize \color{darkred}}
\subsubsectionfont{\normalsize \color{darkred}}
\paragraphfont{\normalsize \color{darkred}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\LARGE
\textbf{Single-cell virus sequencing of influenza infections that trigger innate immunity}}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
{\bf
Alistair B. Russell\textsuperscript{1},
Jacob R. Kowalsky\textsuperscript{2},
Jesse D. Bloom\textsuperscript{1,2,3*}
}
\\
\bigskip
{\small
$^1$Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center
\\
$^2$Department of Genome Sciences, University of Washington
\\
$^3$Howard Hughes Medical Institute
\\
Seattle, WA  98109, USA
\\
*Correspondence: jbloom@fredhutch.org
}
\bigskip

@@

\section*{SUMMARY}
{\bf
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
}

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{INTRODUCTION}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.

@@

\section*{RESULTS}

@@

\section*{DISCUSSION}
We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\citep{saikia2018simultaneous, zanini2018virus}.
To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \citep{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.

@@

\small
\subsection*{ACKNOWLEDGMENTS}
We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
JDB is an Investigator of the Howard Hughes Medical Institute.

\subsection*{AUTHOR CONTRIBUTIONS}
ABR and JDB designed the study.
ABR performed the experiments with assistance from JRK.
ABR and JDB computationally analyzed the data and wrote the paper.

\subsection*{DECLARATION OF INTERESTS}
The authors declare no competing interests.

@@

\section*{METHODS}

@@

\section*{SUPPORTING INFORMATION}
